---
title: "Monocyte analysis V11.1"
author: "Raphael Kübler"
date: "01/05/2021"
output:
  rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true 
---

This analysis consists of four major parts (three of which are essential to its interpretation):
1) the base QC and data correction (based on appropriate model choice)
2) DEA and enrichment analysis (annotation)
3) Dissection of GEX variance in gene sets (delineation of status effect)
4) Networking approach: co-regulations (propagation of status effect on rest of transcriptome)


```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library(xlsx)
library(dplyr)
library(readxl)
library(Hmisc)
library(pheatmap)
library(DESeq2)
library(factoextra)
library(broom)

library(variancePartition)
library(RColorBrewer)
library(pcaExplorer)
library(vsn)

library(clusterProfiler)

library(EnhancedVolcano)
library(plyr)
library(ggpmisc)
library(ggbeeswarm)
library(ggpubr)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")

plot.density <- function(meta.df,to.color="Condition",to.plot="Library size",ylab.title="Density",xlab.title="Library size"){
  meta.df %>% 
    ggplot(aes_string(color=to.color, x=to.plot, fill=to.color)) + 
    geom_density(alpha = 0.2) + 
    scale_x_log10() + 
    theme_classic() +
    ylab(ylab.title) +
    xlab(xlab.title)
}

```

Analysis of monocyte data (SCZ vs CTR): most advanced model

## Load data
```{r load_data, echo=FALSE}
dir <- file.path("D:", "R", "R workspace")
setwd(dir)

load("01052022_Monocyte paper analysis_V11_1.RData")

```

## Base QC

Plotting the first QC results:

(1) Density plot
```{r plots.1, echo=FALSE}
#Density plots
#550 x 450
#Library size; no need to filter: normally distributed
plot.density(imputed.covs,to.color="status",to.plot="`Mapped`",xlab.title="Reads mapped")
plot.density(imputed.covs,to.color="status",to.plot="`Exonic.Rate`",xlab.title="Exonic rate")
plot.density(imputed.covs,to.color="status",to.plot="`Genes.Detected`",xlab.title="Genes detected")
plot.density(imputed.covs,to.color="status",to.plot="`RIN`",xlab.title="RIN")


```

(2) Sample-sample correlation
```{r plots.2, echo=FALSE, fig.height = 6, fig.width = 7.5, fig.align = "center"}
annot_df <- data.frame("Status"=as.character(imputed.covs$status), "Sex"=as.character(imputed.covs$sex), check.names = F)
annot_df$Status[annot_df$Status == 1] <- "SCZ"
annot_df$Status[annot_df$Status == 0] <- "CTR"
annot_df$Sex[annot_df$Sex == 1] <- "Male"
annot_df$Sex[annot_df$Sex == 2] <- "Female"

rownames(annot_df) <- rownames(imputed.covs)
cols <- c("gray","tomato")
cols.2 <- c("#A6CEE3","lightpink")

annot_colors <- list("Status"=cols, "Sex"=cols.2) #annotation colors for heatmap
names(annot_colors$`Status`) <- c("CTR","SCZ")
names(annot_colors$`Sex`) <- c("Male","Female")


ComplexHeatmap::pheatmap(cor(filt.df), annotation_row = annot_df, annotation_col = annot_df,
                         annotation_colors = annot_colors, color = heatmap.color.code)
```

(3) Covariate correlation
```{r norm.qc1, echo=TRUE, fig.height = 10, fig.width = 12, fig.align = "center"}
#Covariate correlation
for (i in colnames(imputed.covs))
  imputed.covs[,i] <- as.numeric(as.character(imputed.covs[,i]))
MEs_lin(imputed.covs,imputed.covs)

```

(4) PC-covariate correlation
```{r QC.pre, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}

#PC-covariate correlation (nomal p)
PCA_cov_cor_R(imputed.covs, filt.df)

```

Next step: QC on vst-normalized expression values.

(1) Inter-sample distances

```{r norm.qc2, echo=FALSE, fig.height = 6, fig.width = 7.5, fig.align = "center"}
ComplexHeatmap::pheatmap(cor(trnsf.df,trnsf.df), annotation_row = annot_df, annotation_col = annot_df,
                         annotation_colors = annot_colors, color = heatmap.color.code)

```

(2) Normalized PCA

```{r qc3.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
for (i in c("status", "smoking", "coffee", "sex", "batch", "drugs", "antipsychotics", "antidepressive", "antianxiety"))
  imputed.covs[,i] <- as.factor(imputed.covs[,i])
pca <- plotPCA.custom(as.matrix(trnsf.df), intgroup=c("status", "batch", "sex", "smoking"), 
                      ntop = 50000, returnData=TRUE, metadata=imputed.covs, pc.1 = 1, pc.2 = 2)
PoV <- round(100 * attr(pca, "percentVar"))
PCAplot(pca, "status", PoV.df=PoV, pc.1 = 1, pc.2 = 2) 

```


(3) PCA post outlier removal
```{r qc3.2, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
PCAplot(pca.clean, "status", PoV.df=PoV.clean, pc.1 = 1, pc.2 = 2)

```


(4) Variance partitioning

```{r qc3.3, echo=TRUE, fig.height = 6, fig.width = 7.5, fig.align = "center"}
#Check influence of covariates on data variance after transformation
plotVarPart(vp)

```


## Data correction
(1) PC-covariate correlation
```{r QC.post, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}
print("The model used is:")
dds@design

#PC-covariate correlation (nominal p)
PCA_cov_cor_R(clean_covs, batch.rem)

```

(2) PCA plots
```{r batchrem.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
PCAplot(pca.cor, "status", PoV.df=PoV.cor, pc.1 = 1, pc.2 = 2)


```

(3) New sample-sample correlations
```{r batchrem.2, echo=FALSE, fig.height = 6, fig.width = 7.5, fig.align = "center"}
annot_df <- annot_df[!rownames(annot_df) %in% c("36", "37"),]
ComplexHeatmap::pheatmap(cor(batch.rem), annotation_row = annot_df, annotation_col = annot_df,
                         annotation_colors = annot_colors, color = heatmap.color.code)
```
## DESeq2 anaylsis

We then go into the differential gene-expression analysis:

(1) Overview of the results 
```{r plots.11, echo=FALSE}
summary(res)
```

As well as the number of differentially expressed genes at lfc </> -/+0.5 at padj < 0.1:

```{r plots.12, echo=FALSE}
sum(res$log2FoldChange < -0.5 & res$padj < 0.1, res$log2FoldChange > 0.5 & res$padj < 0.1, na.rm=TRUE)

```

## Interactive results table

(1) Results

```{r DESeq2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)

createDT(data.frame(res), "Status", scrollY=1000)

```

## Plotting results

(2) Volcano plot:
```{r plots.13, echo=TRUE, fig.height = 6, fig.width = 8, fig.align = "center"}
EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.1,
                FCcutoff = 0.5,
                labSize = 6,
                xlim = c(-2.5,2.5),
                ylim = c(-0.5,2),
                legendPosition = 'right')
```

(3) Plotting top genes (lFC < -1 / > 1) on a heatmap:
```{r plots.14A, echo=TRUE, fig.height = 5, fig.width = 6, fig.align = "center"}

#Plotting DEGs
ComplexHeatmap::pheatmap(batch.rem[rownames(batch.rem) %in% upreg.genesALL,], scale= "row",
                         cluster_rows = T, cluster_cols = T, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)




```

```{r plots.14B, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}

ComplexHeatmap::pheatmap(batch.rem[rownames(batch.rem) %in% downreg.genesALL,], scale= "row",
                         cluster_rows = T, cluster_cols = T, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)

```



## Enrichment testing results

Making meaning of our DEGs

(1) GO enrichments
```{r plots.15, echo=TRUE, fig.height = 9, fig.width = 12, fig.align = "center"}

dotplot(GO.up, split="ONTOLOGY") + facet_grid(ONTOLOGY~., scale="free")

```
```{r B, echo=TRUE, fig.height = 3, fig.width = 10, fig.align = "center"}

print("No GO-term enrichments for downregulated genes")
```


(2) Hand-selected gene set enrichments
```{r Enrich, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)
paste("Reminder: we have", length(upreg.genesALL), "upregulated and", length(downreg.genesALL), "downregulated genes.")

createDT(DEG.enrich.res, "Enrichment", scrollY=1000)

```

```{r plots.15beta, echo=TRUE, fig.height = 3, fig.width = 3, fig.align = "center"}

ComplexHeatmap::pheatmap(DEG.enrich, cluster_rows = F, cluster_cols = F, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(0,-log(0.05), 10, 20, 20), 
                         legend = T, color = colorRampPalette(RColorBrewer::brewer.pal(6,"RdPu"))(30), 
                         legend_labels=c("0","2.995", "10", "20","-log2(q) \n\n\n"))


```


(3) Gene-set overlap
```{r enrich.overlap, fig.height = 5, fig.width = 6, fig.align = "center"}

ComplexHeatmap::pheatmap(overlap.genesets, cluster_rows = F, cluster_cols = F)
```

So we select the NFKB, LPS, and chronic/acute GC stimulation gene sets, as well as the DEX 2.5/5nM stimulation


## Status effect on gene set expression

(1) Gene set DEG expressions
```{r var.1, echo=TRUE, fig.height = 5, fig.width = 6, fig.align = "center"}
gene.sets <- c("NfKB", "LPS", "acute_GC", "chronic_GC", "DEX_2.5nM", "DEX_5nM")

for (i in gene.sets){
  panel <- panel.list[[i]]
  DE <- DE.list[[i]]
  df <- panel[DE$log2FoldChange < -0.5 | DE$log2FoldChange > 0.5,]
  df <- df[,order(as.numeric(colnames(df)), method="radix", decreasing=F)]
  ComplexHeatmap::pheatmap(df, scale= "row",
                           cluster_rows = T, cluster_cols = T, annotation_legend = T, show_colnames = F, show_rownames = T,
                           legend = T, treeheight_row = 0, color = heatmap.color.code,
                           annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8, column_title=paste("DEGs within the", i, "list"))
}

```

(1A) Compound median lFC^2 for each signature
```{r plots.16A, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}
#mean lFC to point directions
compoundlFC <- median(DE.list[[3]][order(DE.list[[3]]$log2FoldChange^2, decreasing=T),][1:10,]$log2FoldChange)
for (i in gene.sets[-1]){
  top10 <- DE.list[[i]][order(DE.list[[i]]$log2FoldChange^2, decreasing=T),][1:10,]
  compoundlFC <- rbind(compoundlFC,median(top10$log2FoldChange))}
rownames(compoundlFC) <- gene.sets


ComplexHeatmap::pheatmap(compoundlFC^2, color = colorRampPalette(RColorBrewer::brewer.pal(6,"RdPu"))(30),
                         cluster_rows = F)


```


(2) Variance causa testing
```{r var.causa, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}

for (i in gene.sets){
 dat <- batch.rem[rownames(batch.rem) %in% na.omit(gene.panels_mrgd[,i]),]
#  plots <- PCA_cov_cor_R(clean_covs, dat) #if 
  list.pca <- plotPCA.custom(as.matrix(dat), intgroup=c("status", "batch", "sex", "RIN"), 
                             ntop = 50000, returnData=TRUE, metadata=clean_covs, pc.1=1, pc.2=2)
  list.pov <- round(100 * attr(list.pca, "percentVar"))
  plot <- PCAplot(list.pca, "status", PoV.df=list.pov, pc.1=1, pc.2=2)
  print(plot)}
```

(2A) Variance causa testing - does GEX variance come really from status?

PC-expression t-test
```{r varPar, echo=TRUE, fig.height = 7, fig.width = 7, fig.align = "center"}
PC.boxplots <- ggarrange(plotlist=scatters[names(scatters) %in% gene.sets], common.legend = T)
PC.boxplots


```



```{r plots.17, echo=TRUE, fig.height = 3, fig.width = 8, fig.align = "center"}
GEX.stat.cor <- GEX.stat.cor[rownames(GEX.stat.cor) %in% gene.sets,]
matrix_pvalue <- matrix_pvalue[rownames(matrix_pvalue) %in% gene.sets,]
ComplexHeatmap::pheatmap(cbind(GEX.stat.cor[,1]), cluster_rows = F, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(-0.6,-0.3,0,0.3,0.6,0.6), 
                         legend = T, color = heatmap.color.code,
                         legend_labels=c("-0.6","-0.3","0","0.3","0.6","Spearman rho \n\n\n"),
                         display_numbers = as.matrix(matrix_pvalue[,1]), column_title="Status-GEX variance correlation")

ComplexHeatmap::pheatmap(cbind(GEX.stat.cor[,3]), cluster_rows = F, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(-3.5,-2.5,-1,0,1,2.5,2.5), 
                         legend = T, color = heatmap.color.code,
                         legend_labels=c("-3.5","-2.5","-1","0","1","2.5","beta \n\n\n"),
                         display_numbers = as.matrix(matrix_pvalue[,2]), column_title="GEX variance prediction based on status (beta effect size)")



```



## NfKB signature co-variances

(1) DEG gene program-NfKB gene program interaction based on PCs
```{r plots.19, echo=TRUE, fig.height = 8, fig.width = 10, fig.align = "center"}
#Compiling a new list of scatter plots
#Compile a list of the usable plots:
DEG.NfKB <- list()
names(DEGscatters$DEG)
for (i in names(DEGscatters)[1:3])
  DEG.NfKB <- append(DEG.NfKB, DEGscatters[[i]]["NfKB"])

#1) co-reg between NfKB and DEGs
DEG.NfKB.compoundScatter <- ggarrange(plotlist=DEG.NfKB[1:length(DEG.NfKB)], common.legend = T)
DEG.NfKB.compoundScatter

```
(2) Gene program-Gene program interaction based on PCs
```{r plots.20, echo=TRUE, fig.height = 8, fig.width = 10, fig.align = "center"}
#2) co-regs between NfKB and all other gene sets

all.NfKB <- DEGscatters[["NfKB"]][!names(DEGscatters[[i]]) %in% c("NfKB","DEG", "DEG.up", "DEG.down")]

all.NfKB.compoundScatter <- ggarrange(plotlist=all.NfKB[1:length(all.NfKB)])
all.NfKB.compoundScatter

```


(3) Data summarized as correlation on heatmap
```{r plots.21, echo=TRUE, fig.height = 6, fig.width = 6.5, fig.align = "center"}

ComplexHeatmap::pheatmap(rcorr(as.matrix(cbind(DEGscatter.df,"Status"=scatter.df[,1])), type="spearman")$r,
                         color = heatmap.color.code, fontsize_row = 8, fontsize_col = 8)
```


(3) One more network question: Does status influence these interactions?
```{r plots.22, echo=TRUE, fig.height = 9, fig.width = 9.5, fig.align = "center"}

ComplexHeatmap::pheatmap(z, cluster_rows = T, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(-2,-1,0,1,2,2), 
                         legend = T, color = heatmap.color.code,
                         legend_labels=c("-2","-1","0","1","2","z-score \n\n\n"),
                         display_numbers = P)
```