---
title: "Monocyte analysis V6"
author: "Raphael Kübler"
date: "11/10/2021"
output:
  rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true 
---

---

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library(xlsx)
library(dplyr)
library(readxl)
library(Hmisc)
library(pheatmap)
library(DESeq2)
library(factoextra)
library(broom)

library(variancePartition)
library(RColorBrewer)
library(pcaExplorer)
library(vsn)

library(clusterProfiler)

library(EnhancedVolcano)
library(plyr)
library(ggpmisc)
library(ggbeeswarm)
library(ggpubr)
knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

Analysis of monocyte data (SCZ vs CTR): simplest model (~ status)

## Load data
```{r load_data, echo=FALSE}
dir <- file.path("D:", "R", "R workspace")
setwd(dir)

load("11102021_monocyte paper analysis_V6.RData")

heatmap.color.code <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(496)

```

## First step QC

Plotting the first QC results:

(1) Density plot
```{r plots.1, echo=FALSE}
#Density plots
#550 x 450
#Library size; no need to filter: normally distributed
hist(log10(int.covs$'Mapped'), xlab="Library sizes (log10 mapped reads)", 
     main=paste("Mapped reads \n\n", "mean = ", round(mean(int.covs$'Mapped'),2), ";",
                "median = ", round(median(int.covs$'Mapped'),2)), 
     breaks=20, col="grey80", ylab="Number of samples")

hist(int.covs$'Exonic.Rate', xlab="Exonic Rate", 
     main=paste("Exonic Rate \n\n", "mean = ", round(mean(int.covs$'Exonic.Rate'),2), ";",
                "median = ", round(median(int.covs$'Exonic.Rate'),2)), 
     breaks=20, col="grey80", ylab="Number of samples")

hist(int.covs$'Genes.Detected', xlab="Genes Detected", 
     main=paste("Genes Detected \n\n", "mean = ", round(mean(int.covs$'Genes.Detected'),2), ";",
                "median = ", round(median(int.covs$'Genes.Detected'),2)), 
     breaks=20, col="grey80", ylab="Number of samples")

hist(int.covs$'RIN', xlab="RIN values", 
     main=paste("RIN value distribution \n\n", "mean = ", round(mean(int.covs$'RIN'),2), ";",
                "median = ", round(median(int.covs$'RIN'),2)), 
     breaks=20, col="grey80", ylab="Number of samples")

```

(2) Covariate correlation
```{r norm.qc1, echo=TRUE, fig.height = 7, fig.width = 7, fig.align = "center"}
#Covariate correlation
for (i in colnames(int.covs))
  int.covs[[i]] <- as.numeric(int.covs[[i]])
MEs_lin(int.covs,int.covs)
```

(3) PC-covariate correlation
```{r QC.pre, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}

#PC-covariate correlation (nomal p)
PCA_cov_cor_R(int.covs,trnsf.df)

```

Next step: QC on vst-normalized expression values.

(1) Gene expression variance (low variance in lowly expressed genes (to the left) = good)

```{r norm.qc2, echo=TRUE, fig.height = 4, fig.width = 6.5, fig.align = "center"}
meanSdPlot(assay(vsd))
```

(2) PCA

```{r qc3.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#Check influence of covariates on data variance after transformation
PCAplot(pca, "batch", PoV.df=PoV, colors=c('#2E7281','black','#50B8CF',"green"), pc.1 = 3, pc.2 = 4)
PCAplot(pca, "status", PoV.df=PoV, pc.1 = 3, pc.2 = 4)

```

## DESeq2 anaylsis

We then go into the differential gene-expression analysis:

(1) Overview of the results 
```{r plots.11, echo=FALSE}
summary(res)
```

As well as the number of differentially expressed genes at lfc </> -0.5/0.5 at padj < 0.1:

```{r plots.12, echo=FALSE}
sum(res$log2FoldChange < -0.5 & res$padj < 0.1, res$log2FoldChange > 0.5 & res$padj < 0.1, na.rm=TRUE)
```

## Interactive results table

(1) Results

```{r DESeq2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)

createDT(data.frame(res), "Status", scrollY=1000)

```

## Plotting results

(2) Volcano plot:
```{r plots.13, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.1,
                FCcutoff = 2,
                labSize = 6,
                xlim = c(-5,5),
                ylim = c(-1,5),
                legendPosition = 'right')
```

(3) Plotting top genes on a heatmap:
```{r plots.14, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}

#Plotting DEGs
ComplexHeatmap::pheatmap(sort.trnsf.df[rownames(sort.trnsf.df) %in% upreg.genes,], scale= "row",
                         cluster_rows = T, cluster_cols = F, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)

ComplexHeatmap::pheatmap(sort.trnsf.df[rownames(sort.trnsf.df) %in% downreg.genes,], scale= "row",
                         cluster_rows = T, cluster_cols = F, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)

```


## Enrichment testing results
(1) List enrichments
```{r Enrich, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)
paste("Reminder: we have", length(upreg.genes), "upregulated and", length(downreg.genes), "downregulated genes.")

createDT(DEG.enrich.res, "Enrichment", scrollY=1000)

```

(2) GO enrichments
```{r plots.15, echo=TRUE, fig.height = 6, fig.width = 17, fig.align = "center"}

dotplot(GO.up, split="ONTOLOGY") + facet_grid(ONTOLOGY~., scale="free")
dotplot(GO.down, split="ONTOLOGY") + facet_grid(ONTOLOGY~., scale="free")
```

## Signature co-variance
(1) Signature expression (Microglia, NfKB, IFNy)
```{r plots.16, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
mic.panel <- mic.panel[,order(int.covs$status, decreasing=F)]
pheatmap::pheatmap(mic.panel, cluster_rows = T, cluster_cols = F, annotation_legend = F, show_colnames = F, show_rownames = T,
                   legend_breaks=c(-4,0,4,4), scale="row",
                   legend = F, treeheight_row = 0,
                   legend_labels=c("-4","0","4","Residual expression \n\n"),
                   annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)
NFKB.panel <- NFKB.panel[,order(int.covs$status, decreasing=F)]

pheatmap::pheatmap(NFKB.panel, cluster_rows = T, cluster_cols = F, annotation_legend = F, show_colnames = F, show_rownames = F,
                   legend_breaks=c(-4,0,4,4), scale="row",
                   legend = F, treeheight_row = 0,
                   legend_labels=c("-4","0","4","Residual expression \n\n"),
                   annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)
IFN.panel <- IFN.panel[,order(int.covs$status, decreasing=F)]

pheatmap::pheatmap(IFN.panel, cluster_rows = T, cluster_cols = F, annotation_legend = F, show_colnames = F, show_rownames = F,
                   legend_breaks=c(-4,0,4,4), scale="row",
                   legend = F, treeheight_row = 0,
                   legend_labels=c("-4","0","4","Residual expression \n\n"),
                   annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)

```

(1A) Signature expression only for lFC < -1 | lfc > 1 genes (Microglia, NfKB, IFNy)
```{r plots.16A, echo=TRUE, fig.height = 6, fig.width = 4, fig.align = "center"}
for (i in c("NFKB", "IFN", "Microglia")){
  panel <- panel.list[[i]]
  DE <- DE.list[[i]]
  df <- panel[DE$log2FoldChange < -0.5 | DE$log2FoldChange > 0.5,]
  df <- df[,order(as.numeric(colnames(df)), method="radix", decreasing=F)]
  annot_df_heat <- data.frame("Status" = int.covs[order(as.numeric(rownames(int.covs), decreasing=F)),]$status, check.names=F)
  rownames(annot_df_heat) <- colnames(df)
  ComplexHeatmap::pheatmap(df, scale= "row",
                         cluster_rows = T, cluster_cols = F, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df_heat, fontsize_row = 8)
}

```

(2) Signature co-variance
```{r plots.17, echo=TRUE, fig.height = 5.5, fig.width = 6.5, fig.align = "center"}
paste("Microglia")
PCA_cov_cor_R(int.covs, mic.panel)
PCAplot(mic.pca, "status", PoV.df=mic.PoV, pc.1=2, pc.2=6)

paste("NFKB")
PCA_cov_cor_R(int.covs, NFKB.panel)
PCAplot(IFN.pca, "status", PoV.df=IFN.PoV, pc.1 = 5, pc.2 = 6)

paste("IFNy")
PCA_cov_cor_R(int.covs, IFN.panel)
PCAplot(NFKB.pca, "status", PoV.df=NFKB.PoV, pc.1=3, pc.2=6)


```

(2A) Scatter plot of signature co-variance
```{r plots.18, echo=TRUE, fig.height = 8, fig.width = 8, fig.align = "center"}
scatters.all

```

(2A) Scatter plot of signature with DEG co-variance (messy, need to clean up)
```{r plots.19, echo=TRUE, fig.height = 8, fig.width = 12, fig.align = "center"}
DEGscatters.all

```


