---
title: "Monocyte analysis V10"
author: "Raphael Kübler"
date: "12/15/2021"
output:
  rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true 
---

---

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library(xlsx)
library(dplyr)
library(readxl)
library(Hmisc)
library(pheatmap)
library(DESeq2)
library(factoextra)
library(broom)

library(variancePartition)
library(RColorBrewer)
library(pcaExplorer)
library(vsn)

library(clusterProfiler)

library(EnhancedVolcano)
library(plyr)
library(ggpmisc)
library(ggbeeswarm)
library(ggpubr)
knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")

plot.density <- function(meta.df,to.color="Condition",to.plot="Library size",ylab.title="Density",xlab.title="Library size"){
  meta.df %>% 
    ggplot(aes_string(color=to.color, x=to.plot, fill=to.color)) + 
    geom_density(alpha = 0.2) + 
    scale_x_log10() + 
    theme_classic() +
    ylab(ylab.title) +
    xlab(xlab.title)
}

```

Analysis of monocyte data (SCZ vs CTR): most advanced model

## Load data
```{r load_data, echo=FALSE}
dir <- file.path("D:", "R", "R workspace")
setwd(dir)

load("12142021_Monocyte paper analaysis_V10.RData")

```

## First step QC

Plotting the first QC results:

(1) Density plot
```{r plots.1, echo=FALSE}
#Density plots
#550 x 450
#Library size; no need to filter: normally distributed
plot.density(imputed.covs,to.color="status",to.plot="`Mapped`",xlab.title="Reads mapped")
plot.density(imputed.covs,to.color="status",to.plot="`Exonic.Rate`",xlab.title="Exonic rate")
plot.density(imputed.covs,to.color="status",to.plot="`Genes.Detected`",xlab.title="Genes detected")
plot.density(imputed.covs,to.color="status",to.plot="`RIN`",xlab.title="RIN")


```

(2) Covariate correlation
```{r norm.qc1, echo=TRUE, fig.height = 10, fig.width = 12, fig.align = "center"}
#Covariate correlation
MEs_lin(new.covs,new.covs)
```

(3) PC-covariate correlation
```{r QC.pre, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}

#PC-covariate correlation (nomal p)
PCA_cov_cor_R(imputed.covs, filt.df)

```

Next step: QC on vst-normalized expression values.

(1) Inter-sample distances

```{r norm.qc2, echo=TRUE, fig.height = 6, fig.width = 6.5, fig.align = "center"}
ComplexHeatmap::pheatmap(cor(filt.df,filt.df), color = heatmap.color.code)
```

(2) PCA

```{r qc3.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#Check influence of covariates on data variance after transformation
pca <- plotPCA.custom(as.matrix(trnsf.df), intgroup=c("status", "batch", "sex", "smoking"), 
                      ntop = 50000, returnData=TRUE, metadata=imputed.covs, pc.1 = 1, pc.2 = 2)
PoV <- round(100 * attr(pca, "percentVar"))
PCAplot(pca, "batch", PoV.df=PoV, colors=c('orange','lightblue','lightgreen',"yellow"), pc.1 = 1, pc.2 = 2)
PCAplot(pca, "status", PoV.df=PoV, pc.1 = 1, pc.2 = 2)



```


(3) Variance partitioning

```{r qc3.2, echo=TRUE, fig.height = 6, fig.width = 7.5, fig.align = "center"}
#Check influence of covariates on data variance after transformation
plotVarPart(vp)

```

## Data correction
(1) PC-covariate correlation
```{r QC.post, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}
print("The model used is:")
dds.2@design

#PC-covariate correlation (nomal p)
PCA_cov_cor_R(imputed.covs, batch.rem)

```

(2) PCA plots
```{r batchrem.1, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
PCAplot(pca.cor, "batch", PoV.df=PoV.cor, colors=c('orange','lightblue','lightgreen',"yellow"), pc.1 = 1, pc.2 = 2)
PCAplot(pca.cor, "status", PoV.df=PoV.cor, pc.1 = 1, pc.2 = 2)


```

## DESeq2 anaylsis

We then go into the differential gene-expression analysis:

(1) Overview of the results 
```{r plots.11, echo=FALSE}
summary(res)
```

As well as the number of differentially expressed genes at lfc </> -/+0.5 at padj < 0.1:

```{r plots.12, echo=FALSE}
sum(res$log2FoldChange < 0 & res$padj < -0.5, res$log2FoldChange > 0.5 & res$padj < 0.1, na.rm=TRUE)

```

## Interactive results table

(1) Results

```{r DESeq2, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)

createDT(data.frame(res), "Status", scrollY=1000)

```

## Plotting results

(2) Volcano plot:
```{r plots.13, echo=TRUE, fig.height = 6, fig.width = 8, fig.align = "center"}
EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.1,
                FCcutoff = 2,
                labSize = 6,
                xlim = c(-5,5),
                ylim = c(-1,5),
                legendPosition = 'right')
```

(3) Plotting top genes (lFC < -1 / > 1) on a heatmap:
```{r plots.14A, echo=TRUE, fig.height = 3, fig.width = 6, fig.align = "center"}

#Plotting DEGs
ComplexHeatmap::pheatmap(batch.rem[rownames(batch.rem) %in% upreg.genes,], scale= "row",
                         cluster_rows = T, cluster_cols = T, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)




```

```{r plots.14B, echo=TRUE, fig.height = 3, fig.width = 6, fig.align = "center"}

ComplexHeatmap::pheatmap(batch.rem[rownames(batch.rem) %in% downreg.genes,], scale= "row",
                         cluster_rows = T, cluster_cols = T, annotation_legend = T, show_colnames = F, show_rownames = T,
                         legend = T, treeheight_row = 0, color = heatmap.color.code,
                         annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)

```

## Enrichment testing results
(1) List enrichments
```{r Enrich, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
#DT::renderDT(data.frame(res.complex), "OH1.5A",scrollY=1000)
paste("Reminder: we have", length(upreg.genes), "upregulated and", length(downreg.genes), "downregulated genes.")

createDT(DEG.enrich.res, "Enrichment", scrollY=1000)

```

```{r plots.15beta, echo=TRUE, fig.height = 3, fig.width = 3, fig.align = "center"}

ComplexHeatmap::pheatmap(DEG.enrich, cluster_rows = F, cluster_cols = F, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(0,-log(0.05), 5, 8, 10), 
                         legend = T, color = colorRampPalette(RColorBrewer::brewer.pal(6,"RdPu"))(30), 
                         legend_labels=c("0","2.995", "5", "8","-log(q) \n\n\n"))


```


(2) GO enrichments
```{r plots.15, echo=TRUE, fig.height = 9, fig.width = 12, fig.align = "center"}

dotplot(GO.up, split="ONTOLOGY") + facet_grid(ONTOLOGY~., scale="free")

```
```{r B, echo=TRUE, fig.height = 3, fig.width = 10, fig.align = "center"}

dotplot(GO.down, split="ONTOLOGY") + facet_grid(ONTOLOGY~., scale="free")
```



## Variance causa testing
(1) Gene list variance disparaties between groups
```{r var.1, echo=TRUE, fig.height = 4, fig.width = 7, fig.align = "center"}
for (i in colnames(gene.panels_mrgd)){
  print(i)
  dat <- batch.rem[rownames(batch.rem) %in% na.omit(gene.panels_mrgd[,i]),]
  plots <- PCA_cov_cor_R(imputed.covs, dat)
  list.pca <- plotPCA.custom(as.matrix(dat), intgroup=c("status", "batch", "sex", "RIN"), 
                             ntop = 50000, returnData=TRUE, metadata=imputed.covs, pc.1=1, pc.2=2)
  list.pov <- round(100 * attr(list.pca, "percentVar"))
  print(PCAplot(list.pca, "status", Shape="batch", PoV.df=list.pov, pc.1=1, pc.2=2))}

```

#```{r var.2, echo=TRUE, fig.height = 4, fig.width = 5.5, fig.align = "center"}
#for (i in colnames(gene.panels_mrgd)){
#  print(i)
#  dat <- batch.rem[rownames(batch.rem) %in% na.omit(gene.panels_mrgd[,i]),]
#  list.pca <- plotPCA.custom(as.matrix(dat), intgroup=c("status", "batch", "sex", "RIN"), 
#                             ntop = 50000, returnData=TRUE, metadata=imputed.covs, pc.1=1, pc.2=2)
#  list.pov <- round(100 * attr(list.pca, "percentVar"))
#  print(PCAplot(list.pca, "status", Shape="batch", PoV.df=list.pov, pc.1=1, pc.2=2))}
#```



## Signature interrogation
(1) Signature expression among most lFC genes (< -0.5 | > 0.5)
```{r plots.16, echo=TRUE, fig.height = 6, fig.width = 6, fig.align = "center"}
for (i in colnames(gene.panels_mrgd)){
  panel <- panel.list[[i]]
  DE <- DE.list[[i]]
  df <- panel[DE$log2FoldChange < -0.5 | DE$log2FoldChange > 0.5,]
  df <- df[,order(as.numeric(colnames(df)), method="radix", decreasing=F)]
  print(i)
  ComplexHeatmap::pheatmap(as.matrix(df), scale= "row",
                           cluster_rows = T, cluster_cols = F, annotation_legend = T, show_colnames = F, show_rownames = T,
                           legend = T, treeheight_row = 0, color = heatmap.color.code,
                           annotation_col=annot_df, annotation_colors = annot_colors, fontsize_row = 8)
}


```

(1A) Compound median lFC for each signature
```{r plots.16A, echo=TRUE, fig.height = 4, fig.width = 4, fig.align = "center"}
#median lFC to point directions
ComplexHeatmap::pheatmap(-compoundlFC, color = heatmap.color.code)


```

(2) GEX-Status interaction
```{r plots.17, echo=TRUE, fig.height = 7, fig.width = 4, fig.align = "center"}
ComplexHeatmap::pheatmap(cbind(GEX.stat.cor[,1]), cluster_rows = T, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(0,0.01,0.05,0.1,0.1), 
                         legend = T, color = colorRampPalette(RColorBrewer::brewer.pal(6,"RdPu"))(30),
                         legend_labels=c("0","0.01","0.05","0.1","Rho \n\n\n"),
                         display_numbers = as.matrix(matrix_pvalue[,1]))

ComplexHeatmap::pheatmap(cbind(GEX.stat.cor[,3]), cluster_rows = T, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(0,1,2.5,2.5), 
                         legend = T, color = colorRampPalette(RColorBrewer::brewer.pal(6,"RdPu"))(30),
                         legend_labels=c("0","1","2.5","beta \n\n\n"),
                         display_numbers = as.matrix(matrix_pvalue[,2]))



```

(2B) As differential PC expression
```{r plots.18, echo=TRUE, fig.height = 8, fig.width = 10, fig.align = "center"}
scatters.all

```


(3) DEG gene program- selected gene program interaction based on PCs
```{r plots.19, echo=TRUE, fig.height = 8, fig.width = 10, fig.align = "center"}
clean.compoundScatters

```
(3A) Gene program-Gene program interaction based on PCs
```{r plots.20, echo=TRUE, fig.height = 8, fig.width = 10, fig.align = "center"}
clean.compoundScatters2

```


(3B) Data summarized as correlation on heatmap
```{r plots.21, echo=TRUE, fig.height = 6, fig.width = 6.5, fig.align = "center"}

ComplexHeatmap::pheatmap(rcorr(as.matrix(cbind(DEGscatter.df,"Status"=scatter.df[,1])), type="spearman")$r,
                         color = heatmap.color.code, fontsize_row = 8, fontsize_col = 8)
```


(4) Does status influence these interactions?
```{r plots.22, echo=TRUE, fig.height = 6, fig.width = 6.5, fig.align = "center"}

ComplexHeatmap::pheatmap(z, cluster_rows = T, cluster_cols = T, annotation_legend = F, show_colnames = T,
                         legend_breaks= c(-2,-1,0,1,2,2), 
                         legend = T, color = heatmap.color.code,
                         legend_labels=c("-2","-1","0","1","2","z-score \n\n\n"),
                         display_numbers = P)
```